Gametic Selection, Sex Ratio Bias, and Transitions Between Sex Determination Systems

### 2017-01-21 LaTeX/Turnover.tex
minor grammar on first reading

\title{Gametic Selection, Sex Ratio Bias, and Transitions Between Sex Determination Systems}
\author{Michael F Scott$*^1$ and Matthew M Osmond$*^2$, and Sarah P Otto$^2$}
\date{}
\maketitle
\noindent
$*$ These authors contributed equally to this work

@@

\begin{abstract}
Sex determination systems are remarkably dynamic; many studied taxa display transitions of sex-determining genes between chromosomes or the evolution of new sex-determining systems. 
%Predominant theories in which new sex-determining systems are favoured by selection involve sex ratio selection or fitness differences between sexes (e.g., sexually antagonistic selection). 
Here, we utilize population genetic models to study the spread of novel sex-determiners in systems with 
%a period of selection during the haploid stage produced by one sex, e.g., pollen or sperm competition. 
haploid gametic selection, e.g., pollen or sperm competition.
Haploid selected loci experience a form of sex-specific selection (because gametic competition occurs predominantly among haploids produced by males) and can cause sex ratios at birth to become biased (because sex ratios are determined by the fertilization success of X- versus Y-bearing pollen/sperm). 
%We find that the evolution of sex determination systems where mothers determine sex at birth (e.g., environmental sex determination where sex is determined at birth) is influenced by classic Fisherian sex ratio selection. {\color{red}(Maybe not true???)}
%However, notably, we find that the spread of new genetic sex determination systems is not affected by sex ratio biases that are caused by gametic selection because sex ratios become biased after parental provisioning has occurred (even if pollen/sperm competition occurs within the mother). 
Notably, we find that the spread of new genetic sex determination systems is not affected by sex ratio biases that are caused by gametic selection because sex ratios become biased after parental provisioning has occurred (even if pollen/sperm competition occurs within the mother). 
In addition, we find that linkage of an ancestral sex chromosome to a locus under haploid selection can favour transitions between male and female heterogamety (e.g., XY to ZW), which is not the case for any forms of diploid sex specific selection (e.g., sexually antagonistic selection).
During these transitions, new sex-determining alleles spread despite breaking up favourable associations that build up between ancestral sex-determining loci and selected loci, reducing population mean fitness. 
Furthermore, a period of selection among haploids can favour the stable maintenance of polymorphic sex determination systems. 
Thus, our models offer several new insights to be explored as information about sex determination in non-model taxa accumulates.

@@

Here, we use mathematical models to find the conditions under which new sex determination systems are favoured by selection when there is a period of selection among haploid gametes/gametophytes. 
Selection among haploid genotypes is thought to occur primarily among pollen/sperm, which can compete whenever there are more pollen/sperm than required for fertilization \citep{Mulcahy:1996ha,JOSEPH:2004haa}. 
Haploid selection may be particularly common in plants, in which 60-70\% of all genes are expressed in the male gametophyte and these genes exhibit stronger signatures of selection than random genes \citep{Borg:2009jpa,Arunkumar:2013iq,Gossmann:2014dua}.
In addition, artificial selection pressures applied to male gametophytes cause the frequency of resistant alleles to increase \citep[e.g.,][]{Hormaza:1996cv,Ravikumar:2003uo,Hedhly:2004iv,Clarke:2004ir}. 
A smaller (but non-negligible) proportion of genes are thought to be expressed and selected in animal sperm, although precise estimates are uncertain \citep{Zheng:2001fi,JOSEPH:2004haa,Vibranovski:2010et}.
\textcolor{red}{add something about meiotic drive here?}

@@

%genetic system
We consider the transition between an ancestral and novel sex determination systems using a three locus model.
Locus \textbf{X} is the ancestral sex-determining region, with alleles $X$ and $Y$ (or $Z$ and $W$).
Locus \textbf{A} is a locus under selection, with alleles $A$ and $a$.
Locus \textbf{M} is a novel sex-determining region, at which the null allele ($M$) is initially fixed in the population such that sex of zygotes is determined by the genotype at the ancestral sex-determining region, \textbf{X} ($XX$ become females and $XY$ become males, or $ZW$ become females and $ZZ$ become males). 
To evaluate the evolution of new sex-determination systems, we consider the invasion, fixation, maintenance, and/or loss of novel sex-determining alleles ($m$) at the \textbf{M} locus. 
%We only do invasion analytically though (is this sentence still ok?)
We assume that the \textbf{M} locus is dominant over the \textbf{X} locus such that zygotes with at least one $m$ allele develop as females with probability $k$ and as males with probability $1-k$, regardless of the \textbf{X} locus genotype.
With $k=0$, the $m$ allele is a masculinizer (i.e., a neo-Y) and with $k=1$ the $m$ allele is a feminizer (i.e., a neo-W).
With intermediate $k$, the $m$ allele confers environmental sex determination such that zygotes develop as females in a proportion ($k$) of the environments they experience. 
Finally, we also analyze a model of maternally-controlled environmental sex-determination (ESD), where mothers with at least one $m$ allele produce daughters with probability $k$.

@@

%I think this should be moved below because we have some 'results' where we don't assume weak selection but we only have the equilibrium calculated analytically for weak selection. 
\subsection*{Resident equilibrium and stability}
In the resident population (allele $M$ fixed), we follow the frequency of $A$ in female gametes (eggs) from an XX female, $p^f_X$, and in X-bearing, $p^m_X$, and Y-bearing, $p^m_Y$, male gametes (sperm). 
We also track the total frequency of Y-bearing male gametes (sperm), $q$, which may deviate from $1/2$ due to meiotic drive in males.

Various forms of selection can maintain a polymorphism at the \textbf{A} locus, including sexually antagonistic selection, overdominance and conflicts between diploid selection and selection upon haploid genotypes \citep[ploidally antagonistic selection,][]{Immler:2012tl} or a combination of these selective regimes. 
Here, we assume that selection and meiotic drive are weak relative to recombination ($s^k$, $t^k$, $\alpha_{\Delta}^k$ of order $\epsilon$). 
The maintenance of a polymorphism at the \textbf{A} locus then requires that

@@

\noindent
which indicates that a polymorphism is maintained under various selective regimes. 
In particular special cases, e.g., no sex-differences in selection or meiotic drive ($s^m=s^f$, $h^m=h^f$, and $\alpha^m=\alpha^f=1/2$), equilibrium allele frequency and stability can be calculated analytically without assuming weak selection. 
%Here, we focus on the case where selection is weak relative to recombination so that we make fewer assumptions about the form of selection.

@@

\noindent
where $V_{A}=\bar{p}(1-\bar{p})$ is the variance in the frequency of $A$ and $D^d=\big{(} \bar{p}s^d+(1-\bar{p})h^d s^d\big{)} -\big{(} \bar{p}h^d s^d+(1-\bar{p}) \big{)}$ corresponds to the difference in fitness between $A$ and $a$ alleles in diploids of sex $d$ ($\bar{p}$ is the leading-order probability of mating with an $A$-bearing gamete from the opposite sex). 
The frequency of Y among male gametes is depends upon the difference in $A$ allele frequency on X- and Y-bearing male gametes and the strength of meiotic drive in favour of the $A$ allele, $q=1/2+\alpha_{\Delta}^m(\hat{p}^m_Y-\hat{p}^m_X)/2+O(\epsilon^3)$.
Without haploid competition or drive ($\alpha_{\Delta}^d=t^d=0$), these results reduce to those of \citet{vanDoorn:2007eu}.

>>>>

\title{Gametic Selection, Sex Ratio Bias, and Transitions Between Sex Determination Systems}
\author{Michael F Scott$^{*1}$ and Matthew M Osmond$^{*2}$, and Sarah P Otto$^2$}
\date{}
\maketitle
\noindent
$*$ These authors contributed equally to this work

@@

\begin{abstract}
Sex determination systems are remarkably dynamic; many studied taxa display transitions between sex-determining genes across chromosomes or the evolution of new sex-determining systems. 
%Predominant theories in which new sex-determining systems are favoured by selection involve sex ratio selection or fitness differences between sexes (e.g., sexually antagonistic selection). 
Here, we utilize population genetic models to study the spread of novel sex-determiners in systems with 
%a period of selection during the haploid stage produced by one sex, e.g., pollen or sperm competition. 
haploid gametic selection, e.g., pollen or sperm competition.
Haploid selected loci experience a form of sex-specific selection when gametic competition occurs predominantly among haploids produced by one sex (e.g.,  pollen competition).
This can cause sex ratios at birth to become biased as sex ratios are determined by the fertilization success of X- versus Y-bearing pollen/sperm (or Z- versus W-bearing eggs). 
%We find that the evolution of sex determination systems where mothers determine sex at birth (e.g., environmental sex determination where sex is determined at birth) is influenced by classic Fisherian sex ratio selection. {\color{red}(Maybe not true???)}
%However, notably, we find that the spread of new genetic sex determination systems is not affected by sex ratio biases that are caused by gametic selection because sex ratios become biased after parental provisioning has occurred (even if pollen/sperm competition occurs within the mother). 
Notably, we find that the spread of new genetic sex determination systems is not affected by sex ratio biases that are caused by gametic selection because sex ratios become biased after parental provisioning has occurred, even if pollen/sperm competition occurs within the mother (e.g., pollen competition in stigma). 
In addition, we find that linkage of an ancestral sex chromosome to a locus under haploid selection can favour transitions between male and female heterogamety (e.g., XY to ZW), which is not the case for any forms of diploid sex specific selection (e.g., sexually antagonistic selection).
During these transitions, new sex-determining alleles spread despite breaking up favourable associations that build up between ancestral sex-determining loci and selected loci, reducing population mean fitness. 
Furthermore, a period of selection among haploids can favour the stable maintenance of polymorphic sex determination systems. 
Thus, our models offer several new insights to be explored as information about sex determination in non-model taxa accumulates.

@@

Here, we use mathematical models to find the conditions under which new sex determination systems are favoured by selection when there is a period of selection among haploid gametes/gametophytes. 
Selection among haploid genotypes is thought to occur primarily among pollen/sperm, which can compete whenever there are more pollen/sperm than required for fertilization \citep{Mulcahy:1996ha,JOSEPH:2004haa}. 
Haploid selection may be particularly common in plants, in which 60-70\% of all genes are expressed in the male gametophyte and these genes exhibit stronger signatures of selection than random genes \citep{Borg:2009jpa,Arunkumar:2013iq,Gossmann:2014dua}.
In addition, artificial selection pressures applied to male gametophytes cause the frequency of resistant alleles to increase \citep[e.g.,][]{Hormaza:1996cv,Ravikumar:2003uo,Hedhly:2004iv,Clarke:2004ir}. 
A smaller (but non-negligible) proportion of genes are thought to be expressed and selected in animal sperm, although precise estimates are uncertain \citep{Zheng:2001fi,JOSEPH:2004haa,Vibranovski:2010et}.
%\textcolor{red}{add something about meiotic drive here?}

@@

%genetic system
We consider the transition between ancestral and novel sex determination systems using a three locus model.
Locus \textbf{X} is the ancestral sex-determining region, with alleles $X$ and $Y$ (or $Z$ and $W$).
Locus \textbf{A} is a locus under selection, with alleles $A$ and $a$.
Locus \textbf{M} is a novel sex-determining region, at which the null allele ($M$) is initially fixed in the population such that sex of zygotes is determined by the genotype at the ancestral sex-determining region, \textbf{X} ($XX$ become females and $XY$ become males, or $ZW$ become females and $ZZ$ become males). 
To evaluate the evolution of new sex-determination systems, we consider the invasion, fixation, maintenance, and/or loss of novel sex-determining alleles ($m$) at the \textbf{M} locus. 
%We only do invasion analytically though (is this sentence still ok?)
We assume that the \textbf{M} locus is dominant over the \textbf{X} locus such that zygotes with at least one $m$ allele develop as females with probability $k$ and as males with probability $1-k$, regardless of the \textbf{X} locus genotype.
With $k=0$, the $m$ allele is a masculinizer (i.e., a neo-Y) and with $k=1$ the $m$ allele is a feminizer (i.e., a neo-W).
With intermediate $k$, the $m$ allele confers environmental sex determination such that zygotes develop as females in a proportion ($k$) of the environments they experience. 
Finally, we also analyze a model of maternally-controlled environmental sex-determination (ESD), where mothers with at least one $m$ allele produce daughters with probability $k$.

@@

%I think this should be moved below because we have some 'results' where we don't assume weak selection but we only have the equilibrium calculated analytically for weak selection. 
\subsection*{Resident equilibrium and stability}
In the resident population (allele $M$ fixed), we follow the frequency of $A$ in female gametes (eggs) from an XX female, $p^f_X$, and in X-bearing, $p^m_X$, and Y-bearing, $p^m_Y$, male gametes (sperm). 
We also track the total frequency of Y-bearing male gametes, $q$, which may deviate from $1/2$ due to meiotic drive in males.

Various forms of selection can maintain a polymorphism at the \textbf{A} locus, including sexually antagonistic selection, overdominance and conflicts between diploid selection and selection upon haploid genotypes \citep[ploidally antagonistic selection,][]{Immler:2012tl} or a combination of these selective regimes. 
In particular special cases, e.g., no sex-differences in selection or meiotic drive ($s^m=s^f$, $h^m=h^f$, and $\alpha^m=\alpha^f=1/2$), equilibrium allele frequency and stability can be calculated analytically without assuming weak selection. 
Here, we assume that selection and meiotic drive are weak relative to recombination ($s^k$, $t^k$, $\alpha_{\Delta}^k$ of order $\epsilon$). 
The maintenance of a polymorphism at the \textbf{A} locus then requires that

@@

\noindent
which indicates that a polymorphism is maintained under various selective regimes. 
%In particular special cases, e.g., no sex-differences in selection or meiotic drive ($s^m=s^f$, $h^m=h^f$, and $\alpha^m=\alpha^f=1/2$), equilibrium allele frequency and stability can be calculated analytically without assuming weak selection. 
%Here, we focus on the case where selection is weak relative to recombination so that we make fewer assumptions about the form of selection.

@@

\noindent
where $V_{A}=\bar{p}(1-\bar{p})$ is the variance in the frequency of $A$ and $D^d=\big{(} \bar{p}s^d+(1-\bar{p})h^d s^d\big{)} -\big{(} \bar{p}h^d s^d+(1-\bar{p}) \big{)}$ corresponds to the difference in fitness between $A$ and $a$ alleles in diploids of sex $d$ ($\bar{p}$ is the leading-order probability of mating with an $A$-bearing gamete from the opposite sex). 
The frequency of Y among male gametes depends upon the difference in $A$ allele frequency on X- and Y-bearing male gametes and the strength of meiotic drive in favour of the $A$ allele, $q=1/2+\alpha_{\Delta}^m(\hat{p}^m_Y-\hat{p}^m_X)/2+O(\epsilon^3)$.
Without haploid competition or drive ($\alpha_{\Delta}^d=t^d=0$), these results reduce to those of \citet{vanDoorn:2007eu}.
